\large Experience Report: End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}

### 2020-04-23 doc/2020-SEFM/paper.tex
Changed 2020-CAV directory to 2020-SEFM

\begin{document}
\lstset{language=Ada}
\lstset{basicstyle=\ttfamily}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\title{End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}}
%
\titlerunning{End-to-End Verification of GR(1) Designs in SPARK}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
%\author{Laura Humphrey\inst{1}\orcidID{0000-1111-2222-3333} \and
%James Hamil\inst{2}\orcidID{1111-2222-3333-4444} \and
%Joffrey Huguet\inst{3}\orcidID{2222--3333-4444-5555}}
\author{Laura Humphrey\inst{1} \and
James Hamil\inst{2} \and
Joffrey Huguet\inst{3}}
%
\authorrunning{L. Humphrey et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
\email{laura.humphrey@us.af.mil} \and
LinQuest Corp., Beavercreek, OH 45431, USA \\
\email{james.hamil.ctr@us.af.mil}  \and
AdaCore, F-75009 Paris, France \\
 \email{huguet@adacore.com}}
%
\maketitle              % typeset the header of the contribution
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 150--250 words.
Manually designing control logic for reactive systems is time-consuming and error-prone. 
An alternative is to automatically generate controllers using ``correct-by-construction'' synthesis approaches. 
Recently, there has been interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
since the required computational complexity is relatively low, and several tools now exist for GR(1) synthesis.
However, while these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
errors in tool implementation can still lead to errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications.
Toward this end we have modified Salty -- a tool that produces executable software implementations of controllers 
from GR(1) specifications in a variety of programming languages -- to produce implementations in SPARK.
SPARK is both a programming language and associated set of verification tools, so it has the potential to enable the ``end-to-end'' verification we desire.
In this paper, we discuss our experience to date using SPARK to implement controllers and verify them against 
a subset of properties comprising GR(1) specifications, namely system initial and transition properties. 
We also discuss lessons learned about how to best encode controllers in SPARK for verification, 
examples in which verification found unexpected controller behaviors, and 
caveats related to the interpretation of GR(1) specifications.

@@

In SPARK, subunits consist of a specification and a body, 
which Salty generates. 
% For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
The package specification for this traffic light example is shown in \figref{fig:trafficLightSpec}.
Type \lstinline{Controller} encodes a Moore machine representing the synthesized controller. 
In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state.  
% (possibly violating the controller's GR(1) specification). 
In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
current input value(s), and the current output value(s). 
Its initial state is always the last possible state number. 
In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
Here, there is only one input, so it is not wrapped in a record. 
Since \lstinline{Controller} is a Moore machine, for each internal state, 
there is exactly one set of output values produced in that state. 
Furthermore, controllers synthesized from GR(1) specifications have the additional property that 
there is exactly one set of input values that brought the controller into that state. 
Therefore, \lstinline{Controller} includes a type invariant (lines 36-37) that captures this property, 
where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} state whether the controller has the corresponding state number and input or output value(s), respectively. 
% They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
Both of these functions are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
unless it is specified with a certain switch during compilation that they should be.
Executing ghost code is often used for debugging, e.g. to check whether an unproved property is actually true. 
In this case, removing the ghost code can save significant memory, 
since \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
internally encode large lookup tables that have to store information on every state of the controller, 
which is often thousands of states or even millions of states for the largest example in our database.
The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
\lstinline{State_To_Input_Mapping (C)} returns \lstinline{True} if and only if \lstinline{C.tick = False} and 
\lstinline{State_To_Output_Mapping (C)} returns \lstinline{True} if and only if 
\sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
since \lstinline{tick = False} brings the system into state 1 and $red$ is true and $yellow$ and $green$ are false in state 1.

>>>>

\begin{document}
\lstset{language=Ada}
\lstset{basicstyle=\ttfamily}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\title{\large Experience Report: End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}}
%
\titlerunning{End-to-End Verification of GR(1) Designs in SPARK}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
%\author{Laura Humphrey\inst{1}\orcidID{0000-1111-2222-3333} \and
%James Hamil\inst{2}\orcidID{1111-2222-3333-4444} \and
%Joffrey Huguet\inst{3}\orcidID{2222--3333-4444-5555}}
\author{Laura Humphrey\inst{1} \and
James Hamil\inst{2} \and
Joffrey Huguet\inst{3}}
%
\authorrunning{L. Humphrey et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
\email{laura.humphrey@us.af.mil} \and
LinQuest Corp., Beavercreek, OH 45431, USA \\
\email{james.hamil.ctr@us.af.mil}  \and
AdaCore, F-75009 Paris, France \\
 \email{huguet@adacore.com}}
%
\maketitle              % typeset the header of the contribution
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 150--250 words.
Manually designing control logic for reactive systems is time-consuming and error-prone. 
An alternative is to automatically generate controllers using ``correct-by-construction'' synthesis approaches. 
Recently, there has been interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
since the required computational complexity is relatively low, and several tools now exist for GR(1) synthesis.
However, while these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
errors in tool implementation can still lead to errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications.
Toward this end we have modified Salty -- a tool that produces executable software implementations of controllers 
from GR(1) specifications in a variety of programming languages -- to produce implementations in SPARK.
SPARK is both a programming language and associated set of verification tools, so it has the potential to enable the ``end-to-end'' verification we desire.
In this paper, we discuss our experience to date using SPARK to implement controllers and verify them against 
a subset of properties comprising GR(1) specifications, namely system initial and transition properties. 
We also discuss lessons learned about how to best encode controllers in SPARK for verification, 
examples in which verification found unexpected controller behaviors, and 
caveats related to the interpretation of GR(1) specifications.

@@

In SPARK, subunits consist of a specification and a body, 
which Salty generates. 
% For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
The package specification for this traffic light example is shown in \figref{fig:trafficLightSpec}.
Type \lstinline{Controller} encodes a Moore machine representing the synthesized controller. 
In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state.  
% (possibly violating the controller's GR(1) specification). 
In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
current input value(s), and the current output value(s). 
Its initial state is always the last possible state number. 
In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
Here, there is only one input, so it is not wrapped in a record. 
Since \lstinline{Controller} is a Moore machine, for each internal state, 
there is exactly one set of output values produced in that state. 
Furthermore, controllers synthesized from GR(1) specifications have the additional property that 
there is exactly one set of input values that brought the controller into that state. 
Therefore, \lstinline{Controller} includes a type invariant (lines 36-37) that captures this property, 
where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} state whether the controller has the corresponding state number and input or output value(s), respectively. 
% They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
Both of these functions are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
unless it is specified with a certain switch during compilation that they should be.
Executing ghost code is often used for debugging, e.g. to check whether an unproved property is actually true. 
In this case, removing the ghost code can save significant memory, 
since \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
internally encode large lookup tables that have to store information on every state of the controller, 
which is often thousands of states or even millions of states for the largest example in our database.
The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
\lstinline{State_To_Input_Mapping(C)} returns \lstinline{True} if and only if \lstinline{C.tick = False} and 
\lstinline{State_To_Output_Mapping(C)} returns \lstinline{True} if and only if 
\sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
since \lstinline{tick = False} brings the system into state 1 and $red$ is true and $yellow$ and $green$ are false in state 1.
